%%%%%%%%%%%%
%
% $Autor: Wings $
% $Datum: 2019-03-05 08:03:15Z $
% $Pfad: AddOns.tex $
% $Version: 4250 $
% !TeX spellcheck = en_GB/de_DE
% !TeX encoding = utf8
% !TeX root = manual 
% !TeX TXS-program:bibliography = txs:///biber
%
%%%%%%%%%%%%

\chapter{Help}

This chapter provides guidance in case you encounter any difficulties while using the Hurricane Intensity Prediction Project.

\section*{Quick Checklist}
Before seeking help, please make sure you have:
\begin{itemize}
	\item Installed all necessary dependencies correctly. (Refer to \textbf{Installation} chapter.)
	\item Followed the initial setup procedures accurately. (Refer to \textbf{First Steps} and \textbf{SetUp} chapters.)
	\item Reviewed the project workflow for better understanding. (Refer to \textbf{Workflow Diagram}.)
\end{itemize}

\section*{Troubleshooting}
If you face technical issues such as errors during execution, incorrect outputs, or missing files, please refer to the \textbf{Troubleshooting} chapter for solutions to common problems.

\section*{Frequently Asked Questions (FAQ)}
\begin{itemize}
	\item \textbf{Q: The program fails to start. What should I check?} \\
	A: Verify your Python version, installed libraries, and dataset availability.
	
	\item \textbf{Q: My model predictions seem off. Why?} \\
	A: Ensure that the preprocessing steps were completed successfully and the dataset was not modified incorrectly.
\end{itemize}

\section*{Further Assistance}
If the provided information does not resolve your issue, consider the following:
\begin{itemize}
	\item Reinstall the environment.
	\item Restart the workflow following the manual carefully.
	\item Look for online tutorials on ARIMA, LSTM, or dataset handling.
\end{itemize}

\section*{Important Reminder}
Since this project runs mainly on Google Colab and uses external libraries (e.g., TensorFlow, Statsmodels), ensure you have a stable internet connection during operations.








