%%%%%%%%%%%%
%
% $Autor: Wings $
% $Datum: 2019-03-05 08:03:15Z $
% $Pfad: AddOns.tex $
% $Version: 4250 $
% !TeX spellcheck = en_GB/de_DE
% !TeX encoding = utf8
% !TeX root = manual 
% !TeX TXS-program:bibliography = txs:///biber
%
%%%%%%%%%%%%

\chapter{Help}

This chapter offers guidance and resources if you encounter any difficulties while using the Hurricane Intensity Prediction System.

\section*{Quick Checklist}
Before seeking additional help, please ensure you have completed the following:
\begin{itemize}
	\item \textbf{Installed all required dependencies.} \\
	Refer to the \textbf{Installation} chapter to confirm all necessary Python packages are present.
	\item \textbf{Followed the initial setup procedures.} \\
	Review the \textbf{First Steps} and \textbf{Setup} chapters to make sure you have properly configured your environment and project files.
	\item \textbf{Understood the project workflow.} \\
	Refer to the \textbf{Workflow Diagram} and relevant sections to familiarize yourself with how data flows through the system.
\end{itemize}

\section*{Troubleshooting}
If you experience technical issues such as errors during execution, missing files, or unexpected results, please consult the \textbf{Troubleshooting} chapter. It provides step-by-step solutions for common problems, including data formatting, model loading, and application errors.

\section*{Frequently Asked Questions (FAQ)}
\begin{itemize}
	\item \textbf{Q: The application does not start. What should I do?} \\
	A: Check that you are using a compatible Python version and that all required libraries are installed. Verify that you are running the correct command (\texttt{streamlit run app.py}) from the project directory.
	
	\item \textbf{Q: My predictions seem inaccurate or unusual. Why?} \\
	A: Ensure your uploaded data is complete and correctly formatted. Double-check that the models in the \texttt{models/} folder are up-to-date and have been trained with appropriate parameters.
	
	\item \textbf{Q: I cannot upload my data file. What could be wrong?} \\
	A: Make sure your file is in the supported format (e.g., CSV) and follows the required structure (date, time, location, wind speed, pressure, etc.). Check for missing or invalid entries.
	
	\item \textbf{Q: The results are not visualized or the download button does not work.} \\
	A: Refresh your browser and ensure you are using a supported web browser. If the issue persists, restart the Streamlit app.
\end{itemize}

\section*{Further Assistance}
If the above steps do not resolve your issue:
\begin{itemize}
	\item Reinstall the required Python packages and dependencies.
	\item Restart the application and carefully follow the user manual steps.
	\item Consult online resources or tutorials for working with ARIMA, LSTM, or data preprocessing in Python.
	\item Contact the developer for advanced support.
\end{itemize}

\section*{Important Reminder}
This Project is designed for research and educational purposes. Always validate your data and review the \textbf{Safety Guidelines} chapter before using model outputs for any important decisions.








