%%%%%%%%%%%%
%
% $Autor: Wings $
% $Datum: 2019-03-05 08:03:15Z $
% $Pfad: Functions.tex $
% $Version: 4250 $
% !TeX spellcheck = en_GB/de_DE
% !TeX encoding = utf8
% !TeX root = manual 
% !TeX TXS-program:bibliography = txs:///biber
%
%%%%%%%%%%%%

\chapter{Installation}

\section{System Requirements}

This project is built to be run easily on local computers. It doesn’t need advanced hardware or cloud services. If your computer is from the last 5–7 years, it’s most likely ready!

\begin{itemize}
	\item \textbf{Computer or Laptop:} Any modern machine (Windows, macOS, or Linux).
	\item \textbf{Mouse and Keyboard:} Required for user interaction and navigation.
	\item \textbf{Monitor/Screen:} At least 1024×768 resolution.
	\item \textbf{Processor:} Dual-core or better (Intel/AMD).
	\item \textbf{RAM:} 4 GB minimum (8 GB+ recommended).
	\item \textbf{Disk Space:} 20–100 MB for project files and data.
	\item \textbf{Internet Access:} Required for installing packages and running the Streamlit app.
	\item \textbf{Python Version:} Python 3.7 or higher (prefer 3.8 or 3.9).
	\item \textbf{Operating System:} Windows 10/11, macOS, Linux.
	\item \textbf{Development Tools:} Jupyter Notebook, JupyterLab, or Streamlit.
	\item \textbf{Web Browser:} Chrome, Firefox, Edge, Safari.
	\item \textbf{Optional – GPU Support:} Accelerates LSTM training but not required.
\end{itemize}


\section{Project Specifications}

\begin{itemize}
	\item \textbf{Training Dataset:} NOAA Atlantic Hurricane Dataset (about 1.36 MB).
	\item \textbf{User Data:} You can upload your own CSV data with hurricane wind speeds.
	\item \textbf{Forecast Models:} ARIMA and LSTM (already trained and ready to use).
	\item \textbf{Programming Language:} Python 3.x (preferably 3.8 or 3.9).
	\item \textbf{Storage:} Very little local space needed. Most work happens in memory.
	\item \textbf{User Interface:} Easy-to-use app using Streamlit.
	\item \textbf{Outputs:} Beautiful graphs and simple tables showing forecast results.
	\item \textbf{Compatibility Note:} TensorFlow doesn't work well with Python 3.10 or newer. Stick to 3.8 or 3.9!
\end{itemize}
