%%%%%%%%%%%%
%
% $Autor: Wings $
% $Datum: 2019-03-05 08:03:15Z $
% $Pfad: SafetyGuidelines.tex $
% $Version: 4250 $
% !TeX spellcheck = en_GB/de_DE
% !TeX encoding = utf8
% !TeX root = manual 
% !TeX TXS-program:bibliography = txs:///biber
%
%%%%%%%%%%%%

\chapter{Safety Guidelines}

When using the Hurricane Intensity Prediction System, please follow these safety guidelines to ensure responsible and secure use of the application and its predictions:

\begin{itemize}
	\item \textbf{Validate Your Data:} Always check that your uploaded hurricane data is accurate, complete, and properly formatted before using it for predictions. Poor data quality can lead to unreliable results and misleading forecasts.
	
	\item \textbf{Understand Model Limitations:} The predictions provided by this system are based on statistical and machine learning models trained on historical data. These models have inherent limitations and uncertainties. Do not rely solely on these predictions for making critical decisions, especially those affecting safety, property, or emergency planning.
	
	\item \textbf{Monitor Model Performance:} Regularly review prediction outputs for unexpected results or errors. If you notice significant discrepancies or unusual forecasts, consult additional trusted sources or contact the system administrator.
	
	\item \textbf{Use Official Forecasts for Safety Decisions:} For real-world hurricane preparedness, always refer to official sources such as the National Hurricane Center (NHC), NOAA, or local authorities for the latest warnings, advisories, and emergency instructions. This system is intended as a supplementary tool for research, analysis, and educational purposes.
	
	\item \textbf{Protect Sensitive Data:} Ensure that any datasets you use or generate, as well as trained model files, are stored securely and are not accessible to unauthorized individuals. Follow your organization’s data privacy and security policies.
	
	\item \textbf{Stay Informed About Hurricane Risks:} Keep up to date with the latest hurricane forecasts and preparedness guidelines from trusted agencies. Enable emergency alerts on your devices and have a plan for responding to severe weather events.
	
	\item \textbf{Do Not Overinterpret Model Outputs:} Remember that all hurricanes are unique, and model predictions may not capture rapid changes or rare events. Use the system’s outputs as one of several sources in your decision-making process.
\end{itemize}








